---
title: "Prediction of DNA methylation"
author: "J Moggridge"
date: "05/04/2021"
output: pdf_document
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = F, warning = F, cache = T)
```

```{r}
library(tidyverse)
library(patchwork)

log_reg_cv <- read_csv("./results/logreg_cv_results.csv") %>% 
  transmute(
    mean_fit_time,
    C = round(param_logistic__C,5),
    penalty = param_logistic__penalty,
    params,
    mean = mean_test_score,
    std_err = std_test_score
  )
log_reg_cv_fig <- log_reg_cv %>% 
  filter(mean > 0.6) %>% 
  ggplot(aes(C, mean, color = penalty, fill = penalty)) +
  geom_ribbon(aes(ymin = mean - std_err, ymax = mean + std_err), 
              color = FALSE, alpha = 0.2) +
  geom_point() +
  geom_path() +
  scale_x_log10() +
  rcartocolor::scale_color_carto_d(palette = 1) +
  rcartocolor::scale_fill_carto_d(palette = 1) +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.8)) +
  labs(y = 'ROC AUC', x = 'regularization C', 
       subtitle = "Logistic regression ")


svm_cv <- read_csv("./results/svm_cv_results.csv")
svm_cv2 <- read_csv("./results/svm2_cv_results.csv") 
svm_cv3 <- read_csv("./results/svm3_cv_results.csv") %>% 
  bind_rows(svm_cv2)%>% 
  transmute(
    mean_fit_time,
    C = round(param_svm__C,5),
    gamma = round(param_svm__gamma, 5),
    params,
    mean = mean_test_score,
    std_err = std_test_score
  ) 
svm_cv_fig <- svm_cv3 %>% 
  filter(mean>0.51) %>% 
  mutate(gamma = as_factor(gamma)) %>% 
  arrange(C) %>% 
  ggplot(aes(x = C, y = mean, color = gamma, fill = gamma)) +
  geom_point(alpha = 0.8, size = 0.75) +
  geom_path(alpha = 0.8) +
  scale_x_log10() +
  scale_color_viridis_d(option = 'C', begin = 0.1, end = 0.92) +
  scale_fill_viridis_d(option = 'C', begin = 0.1, end = 0.92) +
  theme_minimal() +
  labs(y = 'ROC AUC', x = 'regularization (C)', subtitle = 'SVM')

rf_cv <- read_csv("./results/rf_cv_results.csv") %>% 
  transmute(
    estimators = param_rf__n_estimators,
    min_sample_leaf = factor(param_rf__min_samples_leaf),
    max_features = param_rf__max_features,
    mean = mean_test_score,
    std_err = std_test_score
  )
rf_cv_fig <- rf_cv %>% 
  ggplot(aes(x = estimators, y= mean, color = min_sample_leaf,
             fill = min_sample_leaf)) +
    geom_ribbon(aes(ymin = mean - std_err, ymax = mean + std_err), 
              color = FALSE, alpha = 0.2) +
  geom_path() +
  geom_point() +
  scale_x_log10() +
  facet_wrap(~max_features) +
  rcartocolor::scale_color_carto_d(palette = 2) +
  rcartocolor::scale_fill_carto_d(palette = 2) +
  theme_minimal() +
  labs(y = 'ROC AUC', x = 'n estimators', 
       subtitle = "Random forest hyperparameter tuning by 5-fold CV")

rf2_cv <- read_csv("./results/rf2_cv_results.csv") %>% 
  transmute(
    estimators = param_rf__n_estimators,
    min_sample_leaf = factor(param_rf__min_samples_leaf),
    min_samples_split = factor(param_rf__min_samples_split), 
    max_features = param_rf__max_features,
    max_depth = param_rf__max_depth,
    mean = mean_test_score,
    std_err = std_test_score
  )
rf2_cv_fig <- rf2_cv %>% 
  arrange(estimators) %>% 
  ggplot(aes(x = estimators, y= mean, color = min_sample_leaf,
             fill = min_sample_leaf)) +
  geom_point(alpha = 0.85) +
  # scale_x_log10(breaks = c(25, 50, 100, 250)) +
  scale_x_log10() +
  facet_grid(max_features ~ min_samples_split) +
  rcartocolor::scale_color_carto_d(palette = 2) +
  ggthemes::geom_rangeframe() +
  theme_minimal() +
  theme(
    panel.border = element_rect(colour = 'gray', fill = NA),
    panel.grid.minor = element_blank(),
    legend.position = ) +
  labs(y = 'ROC AUC', 
       x = 'n estimators', 
       title = "Random forest hyperparameter tuning by 5-fold CV",
       subtitle = "Panels for max_features (right labels) and min samples per split (top labels)")
rm(svm_cv, svm_cv2, svm_cv3, rf_cv, log_reg_cv)
```

## Introduction

DNA methylation is a heritable, yet reversible epigenetic modification that is involved in regulating transcription and chromatin structure and can be influenced by environmental factors [@Schübeler2015]. Cytosine methylation at specific CpG sites is important for many physiological processes including genomic imprinting, embryonic development, and silencing of transposable elements [@Hollister2009]. Much attention has been directed towards characterizing the role of aberrant DNA methylation patterns in complex disease progression, particularly in cancer [@Vidal2017], but also for rheumatoid arthritis [@cribbs_towards_2015], neuro-degenerative disorders [@nabais_meta-analysis_2021], and others.

Assays to determine the presence or absence of methyl groups at CpG sites, including bisulfite sequencing and hybridization microarrays, have been developed to map methylated sites across the genome. However, these methods can be restricted in their utility due to cost, procedural complexity, limited coverage, and/or high error rates [@kurdyukov2016]. Naturally, predictive models have been developed for this task.

In this work, I sought to develop and evaluate machine learning models for the problem of predicting DNA methylation status at CpG sites. Competitions are frequently used in to compare the efforts of different research groups towards solving particular bioinformatics problems, eg. protein structure prediction (**CITE**). In keeping with this, I used a publicly available dataset of DNA methylation sites that was previously used in a Kaggle competition (<https://www.kaggle.com/c/predict-dna-methylation>) to compare my results with previous efforts. Features were generated using the flanking sequence regions using one-hot-encoding and k-mer counting approaches. For model tuning, I used 5-fold cross validation to evaluate logistic regression, support vector machines (SVM), and random forest (RF) models. The best set of hyperparameters we chosen on the basis of their receiver-operator characteristic area under the curve (AUC). Models were then used to predict the methylation status of unseen CpG sites for comparison.

## Methods

#### Dataset

The DNA methylation dataset used in this work was previously split into a set training of \~30k CpG sites with methylation status and an unlabeled set of 20k for predictions to be submitted for the competition. The data are comprised of positional, categorical, and DNA sequence information. As such, feature extraction must be performed to make use of the non-numeric information, especially the DNA sequences. Categorical predictors include (1) the relation to any nearby CpG island (island, north/south shelf or shore, or none); (2) location relative to genes (TSS regions, UTRs, gene body, 1st exon); (3) associated regulatory features: gene/non-gene/promoter associations & the cell-type specificity of these (T/F).

The positional information include the location within the genome and the position of any nearby CpG islands. The training data is comprised entirely of sites on chromosomes 1-10 and the test data has sites from chromosomes 11-22. I extracted the distance to the nearest island (if one exist) and retained only this information, as the exact location information would be easy for ML models to memorize.

#### Feature extraction

Two popular approaches to DNA sequence processing are to use a 'bag-of-words' approach with k-mers, or to use one-hot encoding for each position. One-hot encoding the sequence has the advantage of retaining positional information whereas the k-mer counts only retain compositional (or motif) information. There is 2 kbp of sequence centered around the CpG sites included in the dataset. I used a combined approach, where I one-hot encoded the 60 bp upstream and downstream regions and counted kmers over the 2 kbp region.

#### Predictive modeling

Machine learning methods were applied to the data using the popular Scikit Learn framework [@scikit-learn]. The labeled data were shuffled and partitioned into stratified subsets for 5-fold cross-validation (20%) and validation (80%). The same folds were used in each cross-validation by setting the random seed. The same cross validation procedure was performed for each of logistic regression, SVM, and RF; to find optimal model tunings a grid-search was perform with a wide range of hyperparameters. For logistic regression, l1 and l2 norms with varied regularization penalty strengths (C); for SVM, separate grid searches were performed with linear and radial basis function (RBF) kernels, where the C and gamma (for RBF only) hyperparameters were tuned. RF models were trained used varied number of trees, minimum samples per split and per leaf, maximum number of features (square root of the number of predictors or 'auto'),

Best models were selected by mean AUC.

Each of these was then used to predict the validation set.

The best model was used to predict the competition test set.

## Results

### Exploratory analysis

```{r fig.width=8.25, fig.height=3.5, fig.cap="Exploratory analysis of DNA methylation data: (top-left) methylation by chromosome; (top-right) presence of regulatory elements, (bottom) relation to CpG island; (right) dinucleotide frequency distribution for methylated and unmethylated CpG sites."}

a <- read_rds("./figures/fig_islands.rds") +
  theme(legend.position = c(.8, .9))
b <- read_rds("./figures/fig_refgene_groups.rds") +
  theme(legend.position = 'none')
c <- read_rds("./figures/fig_chromosomes.rds")  +
  theme(legend.position = 'none')
panel_A <- ((c + b) / (a)) + plot_layout(guides = 'auto')
panel_B <- read_rds("./figures/fig_dinucleotide.rds") +
  theme(legend.position = 'na')
rm(a,b,c)

eda1 <- (panel_A | panel_B )
```

### Exploratory Data Analysis

### Model selection

All model selections and evaluations were performed with the sci-kit learn library.

I first evaluated a basic logistic regression model (l2 norm, C=1.0) to establish a baseline for performance. This model has AUC of 0.923 in cross-validation, which is good, but a surprisingly large AUC of 0.976 on the validation set. If other more complicated models do not perform much better, we would consider this model as it is faster to train and simpler to interpret than SVM, random forest, k-NN, etc., and we can find out which predictors are most important from the model coefficients.

I then tried to improve on the basic logistic regression model by tuning the logistic regression hyperparameters: models had either Lasso or Ridge regularization and penalty (C) values from a geometric distribution over 0.0001 to 1000. The best model in a random search (50 iterations) was a Lasso (l1 norm) model with C = 0.0336. had an AUC of 0.946 in 5-fold CV and 0.985 on the validation set.

To see if a non-linear solution was required to improve the classification performance above 95%, I applied Radial Basis Function (rbf) kernel SVMs in 5-fold cross-validation. The hyperparameters were tuned in a random search over gamma values from ... and C values....

The best model in cross-validation had The ROC AUC was.. On the validation set, the ROC AUC ....

```{r fig.width = 8, fig.height=3, fig.cap="Grid search cross-validation ROC AUC for Logistic regression (left) and SVM (right) classifiers"}
log_reg_cv_fig + svm_cv_fig 
```

```{r}
rf_cv_fig  
rf2_cv_fig
```

```{r}
tribble(
  ~algortithm,   ~roc_auc,  ~accuracy,  ~precision,    ~recall,     ~F1,
  'Random forest',   0.915296,  0.933898,   0.910117,  0.867951,  0.888534,
  'Random forest 2', 0.920104,  0.937812,    0.91632,  0.875035,  0.895202,
'Logistic regression',  0.98546,  0.950026,   0.908537,  0.928875,  0.918593,
  'SVM',             0.932189,  0.941897,   0.901732,  0.907481,  0.904597,
  'SVM3',            0.935965,  0.945209,   0.907554,   0.91244,   0.90999,
)
```

### Model evaluation

## Discussion

    91 min
    RandomizedSearchCV(
      cv=5,
      estimator=Pipeline(steps=[('scaler', StandardScaler()),
      ('svm', SVC(random_state=0))]),
      n_iter=40, n_jobs=-1,
      param_distributions={
        'svm__C': array([
          1e-02, 4.64e-02, 2.1e-01, 1e+00, 4.6e+00, 2.1e+01, 1+02, 4.6e+02,
          2.15e+03, 1e+04
          ]),
      'svm__gamma': array([
          1e-03, 4.64e-03, 2.15e-02, 1e-01, 4.64e-01, 2.1e+00, 1e+01, 4e01,
          202, 1e+03
          ])},
      scoring=make_scorer(roc_auc_score), verbose=3)

    * it used the smallest gamma value and medium in C
    Best SVM score (CV score=0.933):
    Best SVM parameters: {'svm__gamma': 0.001, 'svm__C': 1.0}

                  precision    recall  f1-score   support

               0       0.96      0.96      0.96     16194
               1       0.90      0.91      0.90      7058

        accuracy                           0.94     23252
       macro avg       0.93      0.93      0.93     23252
    weighted avg       0.94      0.94      0.94     23252

    [[15496   698]
     [  653  6405]]

    Test performance of SVM model
      algortithm   roc_auc  accuracy  precision    recall        F1
    0        SVM  0.932189  0.941897   0.901732  0.907481  0.904597



    Best svm3 score (CV score=0.934)
    Best svm3 parameters:
    {'svm__C': 12.915496650148826, 'svm__gamma': 0.0013894954943731374}

                  precision    recall  f1-score   support

               0       0.96      0.96      0.96     16194
               1       0.91      0.91      0.91      7058

        accuracy                           0.95     23252
       macro avg       0.93      0.94      0.94     23252
    weighted avg       0.95      0.95      0.95     23252

    [[15538   656]
     [  618  6440]]

    Test performance of svm3 model

      algortithm   roc_auc  accuracy  precision   recall       F1
    0       SVM3  0.935965  0.945209   0.907554  0.91244  0.90999



    Best RF score (CV score=0.915):
    Best RF params
    {'rf__n_estimators': 200, 'rf__min_samples_leaf': 10, 
    'rf__max_features': 'auto', 'rf__bootstrap': True}


                  precision    recall  f1-score   support

               0       0.94      0.96      0.95     16194
               1       0.91      0.87      0.89      7058

        accuracy                           0.93     23252
       macro avg       0.93      0.92      0.92     23252
    weighted avg       0.93      0.93      0.93     23252

    [[15589   605]
     [  932  6126]]

          algortithm   roc_auc  accuracy  precision    recall        F1
    0  Random forest  0.915296  0.933898   0.910117  0.867951  0.888534

    Best RF2 score (CV score=0.924):
    {'rf__n_estimators': 50, 'rf__min_samples_split': 20, 'rf__min_samples_leaf': 1, 'rf__max_features': 'auto', 'rf__max_depth': 100}
                  precision    recall  f1-score   support

               0       0.95      0.97      0.96     16194
               1       0.92      0.88      0.90      7058

        accuracy                           0.94     23252
       macro avg       0.93      0.92      0.93     23252
    weighted avg       0.94      0.94      0.94     23252

    [[15630   564]
     [  882  6176]]

            algortithm   roc_auc  accuracy  precision    recall        F1
    0  Random forest 2  0.920104  0.937812    0.91632  0.875035  0.895202

    Best score (ROC AUC=0.946)
    'logistic__solver': 'liblinear', 
    'logistic__penalty': 'l1', 
    'logistic__C': 0.03359818286283781

                  precision    recall  f1-score   support

               0       0.97      0.96      0.96     16194
               1       0.91      0.93      0.92      7058

        accuracy                           0.95     23252
       macro avg       0.94      0.94      0.94     23252
    weighted avg       0.95      0.95      0.95     23252


    [[15534   660]
     [  502  6556]]

    Test performance of Logistic regression model   
              algortithm  roc_auc  accuracy  precision    recall        F1
     Logistic regression  0.98546  0.950026   0.908537  0.928875  0.918593

<!-- Beta 0= unmethylated, Beta 1=methylated) -->

## References
